Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-05-22 jss/cheem_paper.Rmd
-fix app link -minor lint&wordsmithing -motiviating letter dates

---
documentclass: jss
author:
  - name: Nicholas Spyrison
    orcid: 0000-0002-8417-0212
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Dept of Human Centred Computing
      | Faculty of Information Technology
    email: \email{spyrison@gmail.com}
    url: https://nspyrison.netlify.app
  - name: Dianne Cook
    orcid: 0000-0002-3813-7155
    address: |
      | Dept of Econometrics and Business Statistics
    affiliation: Monash University
  - name: Przemyslaw Biecek
    orcid: 0000-0001-8423-1823
    affiliation: |
      | University of Warsaw
      | Warsaw University of Technology
title:
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain:     "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  # For running headers, if needed
  short:     "Exploring Local Explanations"
abstract: >
    The increased predictive power of machine learning models comes at the cost of increased complexity and loss of interpretability, particularly in comparison to parametric statistical models. This trade-off has led to the emergence of eXplainable AI (XAI) which provides methods, such as local explanations (LEs) and local variable attributions (LVAs), to shed light on how a model use predictors to arrive at a prediction. These provide a point estimate of the linear variable importance in the vicinity of a single observation. However, LVAs tend not to effectively handle association between predictors. To understand how the interaction between predictors affects the variable importance estimate, we can convert LVAs into linear projections and use the radial tour. This is also useful for learning how a model has made a mistake, or the effect of outliers, or clustering of observations.  The approach is llustrated with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain:     [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: paper.bib
---

@@

- A version of the cheem viewer shiny app can be directly accessed at
https://ebsmonash.shinyapps.io/cheem_initial/.
- The development version of the package is available at https://github.com/nspyrison/cheem, and
- Documentation of the package can be found at https://nspyrison.github.io/cheem/.

>>>>

---
documentclass: jss
author:
  - name: Nicholas Spyrison
    orcid: 0000-0002-8417-0212
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Dept of Human Centred Computing
      | Faculty of Information Technology
    email: \email{spyrison@gmail.com}
    url: https://nspyrison.netlify.app
  - name: Dianne Cook
    orcid: 0000-0002-3813-7155
    address: |
      | Dept of Econometrics and Business Statistics
    affiliation: Monash University
  - name: Przemyslaw Biecek
    orcid: 0000-0001-8423-1823
    affiliation: |
      | University of Warsaw
      | Warsaw University of Technology
title:
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain:     "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  # For running headers, if needed
  short:     "Exploring Local Explanations"
abstract: >
    The increased predictive power of machine learning models comes at the cost of increased complexity and loss of interpretability, particularly in comparison to parametric statistical models. This trade-off has led to the emergence of eXplainable AI (XAI) which provides methods, such as local explanations (LEs) and local variable attributions (LVAs), to shed light on how a model use predictors to arrive at a prediction. These provide a point estimate of the linear variable importance in the vicinity of a single observation. However, LVAs tend not to effectively handle association between predictors. To understand how the interaction between predictors affects the variable importance estimate, we can convert LVAs into linear projections and use the radial tour. This is also useful for learning how a model has made a mistake, or the effect of outliers, or clustering of observations. The approach is llustrated with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain:     [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
preamble: >
  \usepackage{amsmath}
  \usepackage{datetime}
output: rticles::jss_article
bibliography: paper.bib
---

@@

- A version of the cheem viewer shiny app can be directly accessed at
https://ebsmonash.shinyapps.io/cheem/.
- The development version of the package is available at https://github.com/nspyrison/cheem, and
- Documentation of the package can be found at https://nspyrison.github.io/cheem/.

### 2023-05-22 paper.Rmd
-fix app link -minor lint&wordsmithing -motiviating letter dates

---
title: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
type: ARTICLE TEMPLATE
author:
  - name: Nicholas Spyrison
    affiliation: a
    orcid: 0000-0002-8417-0212
    email: spyrison@gmail.com
  - name: Dianne Cook
    affiliation: a
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
  - name: Przemyslaw Biecek
    affiliation: b
    orcid: 0000-0001-8423-1823
    email: przemyslaw.biecek@gmail.com
affiliation:
  - num: a
    address: |
      Monash University, Melbourne, Australia
  - num: b
    address: |
      Warsaw University of Technology and University, Warsaw, Poland
abstract: |
  The increased predictive power of nonlinear models comes at the cost of interpretability of its terms. This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction.  "Local explanations", which provide a point estimate of the linear variable importance in the vicinity of one observation, are one XAI method. These can be considered to be linear projections and can be further explored to understand the interactions between variables used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation and illustrate with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the __R__ package __cheem__, available on CRAN.
keywords: | 
  explainable artificial intelligence; nonlinear model interpretability; visual analytics; local explanations; grand tour; radial tour
bibliography: paper.bib
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

## chunk options
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = FALSE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = FALSE,
  cache.lazy = FALSE,
  fig.height = 5,
  fig.width = 8,
  out.width = "100%"
)
```
<!-- bookdown::pdf_document2: allows \@ref(), pdf_document: requires \ref{} -->
<!-- # cheat sheet {#sec:cheatsheet} -->
<!-- A bib reference [@wickham_visualizing_2015]. -->
<!-- A [Section intro](#sec:cheatsheet) reference, alternatively, section \@ref(sec:intro) (with @; \\@ref(sec:intro)). -->
<!-- ```{r crest, echo=FALSE, out.height = "10%", out.width = "10%", fig.cap = "A caption for crest figure"} -->
<!-- knitr::include_graphics("./figures/crest.jpg") -->
<!-- ``` -->
<!-- A figure \@ref(fig:crest) reference (with @; \\@ref(fig:crest). -->
<!-- (ref:baseplotcap) Separate caption created above the __R__ chunk -->
<!-- ```{r baseplot, echo=F, fig.cap = "(ref:baseplotcap)"} -->
<!-- plot(1,2) -->
<!-- ``` -->

>>>>

---
documentclass: jss
title: 
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  short: "Exploring Local Explanations"
author:
  - name: Nicholas Spyrison
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      Monash University
    email: \email{spyrison@gmail.com}
  - name: Dianne Cook
    affiliation: Monash University
  - name: Przemyslaw Biecek
    affiliation: Warsaw University of Technology and University
abstract: >
  The increased predictive power of nonlinear models comes at the cost of interpretability of its terms. This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction.  "Local explanations", which provide a point estimate of the linear variable importance in the vicinity of one observation, are one XAI method. These can be considered to be linear projections and can be further explored to understand the interactions between variables used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation and illustrate with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the __R__ package __cheem__, available on CRAN.
keywords:  
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
bibliography: paper.bib
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{datetime}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
cls: jss.cls
---

@@

## chunk options
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = FALSE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = FALSE,
  cache.lazy = FALSE,
  fig.height = 5,
  fig.width = 8,
  out.width = "100%"
)
options(tinytex.clean = FALSE)
options(tinytex.verbose = TRUE)
```
<!-- bookdown::pdf_document2: allows \@ref(), pdf_document: requires \ref{} -->
<!-- # cheat sheet {#sec:cheatsheet} -->
<!-- A bib reference [@wickham_visualizing_2015]. -->
<!-- A [Section intro](#sec:cheatsheet) reference, alternatively, section \@ref(sec:intro) (with @; \\@ref(sec:intro)). -->
<!-- ```{r crest, echo=FALSE, out.height = "10%", out.width = "10%", fig.cap = "A caption for crest figure"} -->
<!-- knitr::include_graphics("./figures/crest.jpg") -->
<!-- ``` -->
<!-- A figure \@ref(fig:crest) reference (with @; \\@ref(fig:crest). -->
<!-- (ref:baseplotcap) Separate caption created above the __R__ chunk -->
<!-- ```{r baseplot, echo=F, fig.cap = "(ref:baseplotcap)"} -->
<!-- plot(1,2) -->
<!-- ``` -->
